Chaotic mixing of tidal debris

### 2015-06-24 ms.tex
incorporating comments from kvj and sp

In order to explore the question of how chaos manifests over timescales much shorter than that predicted from generic chaos indicators, we must first understand the structure and behavior of orbits in complex gravitational potentials. An orbit in an $N$ degree of freedom (dof) Hamiltonian, $H$, is a set of $2N$ quasi-periodic time series, 
\begin{equation}
(w_1(t),...,w_{2N}(t)) = (q_1(t),...,q_{N}(t),p_1(t),...,p_{N}(t)) \label{eq:coords}
\end{equation}
where $q_i$ and $p_i$ are conjugate coordinates in the sense that
\begin{align}
	\dot{p}_i &= -\frac{\partial H}{\partial q_i}\\
	\dot{q}_i &= \frac{\partial H}{\partial p_i}
\end{align}
for all $t$. If bounded, the motion in any component, $w_i(t)$, can be represented as a Fourier sum,
\begin{equation}
	w_i(t) = \sum_k^\infty a_k \, e^{i\,\omega_k\,t} \label{eq:fourier}
\end{equation}
where the $a_k$ are complex amplitudes.

@@

A Hamiltonian or potential is said to be \emph{globally integrable} when the number of isolating integrals of motion is equal to the number of degrees of freedom and a transformation to angle-action coordinates may be done globally --- for example, the transformation to angle-action coordinates may be written as a function of arbitrary phase-space coordinates and the functional form is independent of phase-space position. The condition for global integrability is very restrictive and it seems that the likelihood that a Hamiltonian is globally integrable decreases as the number of degrees of freedom increase \citep[e.g.,][]{lichtenberg83}. In a globally integrable potential, the Hamiltonian may be written solely in terms of the actions, $H = H(\boldsymbol{J})$. Galactic potentials are almost certainly not globally integrable, but it is useful to understand the orbit structure in integrable systems before extending to more general potentials. Only one globally integrable triaxial mass model is known that generates motion that is qualitatively similar to that seen in Galaxies: the St\"ackel or `Perfect Ellipsoid' model \citep[e.g.,][]{kuzmin73, deZeeuw85}. The four general classes of orbits in this potential are box, inner long-axis tube, outer long-axis tube, and short-axis tube orbits. The tube orbits preserve a sense of rotation about either the long or short axis and are therefore centrophobic, whereas the box orbits may approach arbitrarily close to the center of the potential (centrophilic) and may change sense of rotation about any axis.

The frequencies of a generic, non-resonant orbit are incommensurable --- that is, $\bs{n} \cdot \bs{\Omega} \neq 0$ for any integer vector, $\bs{n}$, with reasonable magnitude.\footnote{A more precise condition is stated in terms of a diophantine condition, e.g., $|\bs{n} \cdot \boldsymbol{\Omega}| > \alpha \, |n|^{-\gamma}$ where $\alpha, \gamma>0$.} These conditionally-periodic orbits uniformly cover the surface of an orbital torus. If instead there exists a relation of the form $\boldsymbol{n} \cdot \boldsymbol{\Omega} = 0$ the orbit is a resonant orbit. For resonant orbits, the existence of an additional \emph{local} constant of motion reduces the dimensionality of the orbit. We refer to orbits that obey a single resonance relation as \emph{uni-resonant} orbits, and if an additional resonance relation exists, \emph{bi-resonant}. The resonant structure of a potential --- the relative importance of particular resonance numbers --- is difficult to compute, but determines the global behavior of orbits in the potential. For a triaxial potential the resonances define planes in frequency-space, but resonance surfaces in action-space are generic surfaces. In plots of frequency ratios, the resonances appear as lines; Figure~\ref{fig:cartoons} (left panel) shows a cartoon portrait of a portion of frequency-space for an integrable potential with example resonance lines, non-resonant, and resonant orbits marked. For an integrable potential, all non-resonant orbits are regular.

@@

The orbit structure of near-integrable potentials can be understood by considering a Hamiltonian that is a small perturbation away from being globally integrable --- that is, a Hamiltonian that may be written
\begin{equation}
	H(\boldsymbol{J}, \boldsymbol{\theta}) = H_0(\boldsymbol{J}) + \epsilon \, H_1(\boldsymbol{J}, \boldsymbol{\theta})
\end{equation}
where $H_0(\bs{J})$ represents an integrable Hamiltonian, and $\epsilon$ is a small parameter that determines the perturbation strength \citep[a description of perturbation theory applied to nonlinear Hamiltonians is given in][]{lichtenberg83}. When $0 < |\epsilon| \ll 1$, resonant surfaces become `thick' resonant layers, within which orbits are qualitatively similar to the parent resonant orbit \citep[e.g.,][]{merritt99}. These resonant layers are then generically surrounded by stochastic layers where chaotic motion occurs. Chaotic orbits are characterized by exhibiting random behavior despite being generated by entirely deterministic equations of motion. Stochastic layers form gaps in action space where tori are destroyed; the orbits that exist in these regions therefore can not be represented in angle-action coordinates, as their actions and frequencies are not constant and diffuse chaotically with time.

Figure~\ref{fig:cartoons} (middle panel) shows a cartoon of frequency space for a near-integrable potential (a small perturbation away from the potential in the left panel). Much of the structure that was present in the integrable potential remains in the near-integrable case, but the differences are highlighted. Orbits in the resonant layers surrounding the resonances (grey) are near-resonant orbits that will librate around the resonance. Chaotic orbits in the stochastic layer (red) behave erratically depending on the surrounding resonance structure. If the stochastic layer is small and the chaotic orbit is therefore confined, the orbit may behave nearly regular for long periods of time.

For small values of $\epsilon$, many regular orbits survive and only small chaotic regions are introduced, especially where resonances overlap. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Figure~\ref{fig:cartoons} (right panel) qualitatively shows this phenomenon --- the resonant and resonant-layer orbits may still be regular, but many or all of the conditionally-periodic orbits can be chaotic. As the perturbation strength increases, next the uni-resonant tori are destroyed, and finally the bi-resonant tori --- these are least susceptible to destruction from perturbations \cite[for a more quantitative illustration of this transition from integrability to global chaos, see Figure~9 in][]{valluri98}.

When $\epsilon$ is large, there is no general prediction for the resulting behavior, however it seems that more complicated and physically motivated triaxial potential models for galaxies follow the intuition gained from the small-perturbation picture described above, at least for certain parameter choices \citep[e.g.,][]{valluri98, merritt99}. We therefore expect a large number of regular orbits will survive --- the so-called KAM tori --- however the tori that survive will be separated by regions of chaotic motion. Any transformations to angle-action coordinates must be defined locally due to the destruction of tori due to chaotic motion.

@@

Chaotic orbits have no orbital actions because they are only strictly bound to their energy hypersurface (if the potential is time-independent). The orbits therefore do not have a single set of fundamental frequencies, but rather the frequencies that describe the character of motion evolve with time. In near-integrable potentials, chaotic diffusion of an orbit in actions and frequencies occurs both across resonance layers (a sort of stochastic libration) and along the stochastic layers that surround the resonances (Arnold diffusion). For weakly chaotic orbits, motion across a resonance layer can occur with a frequency close to the libration frequency of the nearby stable orbits in the resonance layer. Thus, if the resonance libration frequency is comparable (within a factor of a few) to the orbital frequency, motion across a resonance can modulate the frequency spectrum of an orbit over an orbital time. However, the stochastic layers are often bounded in the direction orthogonal to the resonance by other stable, resonant regions so that the frequencies or actions can't change by large factors (unless there is resonance overlap and the motion is strongly chaotic).

The rate of diffusion along stochastic layers via Arnold diffusion depends on the local resonant structure and is hard to predict. This has been done analytically for simple potentials \citep[e.g.,][]{chirikov??}. For systems with $N>2$ dof, the stochastic layers connect and form an intricate network of stochasticity known as the Arnold web; an orbit that ergodically mixes over its energy hypersurface must traverse this web, though the timescales typical for this phenomenon are many thousands of orbital periods.

For galactic timescales, Arnold diffusion is not expected to be relevant, however chaotic motion across resonances can occur over short times. If a stochastic orbit is semi-bounded by surrounding resonances, the allowed volume in which it can explore (in frequency space) is often quite small so that the orbit may appear to be approximately regular. However, this small-scale evolution may be important for tidal debris which has small intrinsic spreads in frequencies.

\subsection{Mixing of orbit ensembles}

Ensembles of orbits stuck between resonances may quickly spread to fill the allowed volume via the small-scale but short-time chaotic diffusion processes discussed above. This occurs due to phase-mixing of the fundamental frequencies of the ensemble orbits themselves --- the frequencies are evolving and thus trace out an orbit through frequency space. This is in contrast to phase-mixing of regular orbits and long-term chaotic mixing of strongly chaotic orbits.

A small ensemble of (conditionally-periodic) regular orbits will reach a fully phase-mixed state in a timescale  $\approx\sigma_\Omega^{-1}$, where $\sigma_\Omega$ is the dispersion in frequency-space of the ensemble --- each individual orbit will fill the surface of its torus ergodically, but the ensemble will only mix due to sheering from slight variations in their fundamental frequencies \citep{merritt96, helmi99}. Ensembles of chaotic orbits behave quite differently. Over long times, a chaotic orbit in an autonomous Hamiltonian will ergodically mix over the energy hypersurface (a 5D surface, compared to the 3D stable tori). Generically, a small ensemble of chaotic orbits will lose coherence much faster than for regular orbits, however analytically computing the rate and amount of spreading is very difficult even for simple potentials \cite[see, e.g.,][]{chirikov??}. Typically these rates are comparable or larger than the Arnold diffusion timescale and are thus very long.

In this work, we investigate the effect of the short-time but small-scale frequency evolution along weakly chaotic orbits and discuss how this effects our understanding of the coherence and density evolution of tidal streams.

@@

Our goal is to map the orbit structure of arbitrary potentials, with an emphasis on identifying the chaotic orbits and understanding the evolution of these orbits over short times. In this section, we describe the methods we will use to detect and quantify the strength of chaos for large grids of orbits.

@@

\begin{table*}[ht]
\begin{center}
	\begin{tabular}{ c  c }
	         Parameter & Value\\\toprule
		$v_c$ & 175~km~s$^{-1}$\\
		$r_s$ & 20~kpc\\
		$a$ & 1\\
		$b$ & 0.77\\
		$c$ & 0.55\\
		\bottomrule
		\end{tabular}
	\caption{Summary of potential parameters for the triaxial NFW potential (Equation~\ref{eq:potential}) used in this work. Triaxiality is introduced in the density to ensure the density is physical at all radii. Velocity scale, scale radius, and axis ratios are chosen to match the median halo parameters for a $M_{\rm vir} \approx 10^{12}~\msun$ halo from \citep{jing02}. \label{tbl:potential}}
\end{center}
\end{table*}

@@

If an orbit is chaotic, the motion can no longer be expanded in terms of a single set of fundamental frequencies. For a weakly chaotic orbit, the orbit may appear consistently periodic over long windows of time. \superfreq\ will pick out a set of frequencies for such orbits that correspond to the largest peaks in the power spectrum of the stochastic orbits, however these peaks will change location and amplitude with time. For more strongly chaotic orbits, the power spectrum will be quite noisy and the peak frequencies may change erratically when comparing two consecutive sections of orbit. The frequencies picked out by \superfreq\ for such orbits will therefore represent the average periodic nature of the orbit over a given integration window. We define the fractional frequency diffusion rate per orbital period, $\mathcal{R}$, in the $k$th fundamental frequency as
\begin{equation}
	\mathcal{R}_k = \frac{\Omega_{k}^{(2)} - \Omega_{k}^{(1)}}{\Omega_{k}^{(1)} \, N} \label{eq:fdrate}
\end{equation}
where the upper index refers to the two consecutive sections of orbit and $N$ is the number of orbital periods in a single section of orbit. By inverting this rate, we can compute the timescale over which we expect order-unity changes to the fundamental frequencies: the \emph{frequency diffusion time} is defined as
\begin{equation}
	t_\Omega = \, (\max_{a_k} \, \mathcal{R}_k)^{-1} \label{eq:fdtime}
\end{equation}
where the maximum is taken with respect to the corresponding amplitudes, $a_k$, of the fundamental frequency components \citep[see][]{valluri12}, and the time is in units of number of orbital periods, ${\rm T}$.

@@

In Section~\ref{sec:results1}, we generate grids of orbits in the potential described above to map the orbit structure of the potential. We classify each orbit in terms of the strength of chaos along the orbit as computed using the methods of Section~\ref{sec:methods}. With this initial classification, in Section~\ref{sec:results2} we follow the evolution of ensembles of orbits generated around each orbit in the initial grid. We find that the configuration-space density of ensembles around weakly chaotic orbits evolve faster (e.g., in mean density) than expected given the timescales over which chaos is computed to be relevant for the parent orbits. In Section~\ref{sec:results3}, we explain this phenomenon in the context of how chaotic diffusion occurs (e.g., Section~\ref{sec:behavior-chaotic}).

@@

We generate an isoenergy grid of initial conditions along the $xz$ ($y=0$) plane\footnote{$x$ is the major and $z$ the minor axis.} with energy (per unit mass), $E$, chosen to span a range of distances comparable to the scale radius of the potential ($E=-(397.2~{\rm km~s}^{-1})^2$ in physical units; see Table~\ref{tbl:potential}). We fix $v_x = v_z = 0$, and compute $v_y$ from the energy. The majority of the orbits generated on this grid are tube orbits, which preserve a sense of rotation about one of the principal axes of the potential and are generally centrophobic. Tube orbits are generally less stochastic than box orbits, which tend to plunge deep into the inner regions of the potential where the density profile is steep, however there are appreciable numbers of stochastic tube orbits. The major classes of stable tube orbits circulate about either the major or minor axis. This grid generates all of the major orbit classes (short-axis tubes, inner long-axis tubes, outer long-axis tubes, stochastic intermediate-axis, and box orbits), but the most numerous orbits in this grid are the short-axis and outer long-axis tubes. Thin tidal streams are thought to form along tube orbits rather than box orbits because of the faster disruption expected for stellar systems on radially plunging orbits.

@@

For each orbit, we also separately integrate for $\approx$80 orbital periods and use \superfreq\ to compute the fundamental frequencies for the two consecutive sections of 40 orbital periods. We have found that this is the shortest integration window for which we are able to consistently and accurately recover the frequencies using \superfreq. Figure~\ref{fig:freqdiff} shows the same grid of initial conditions, but now the greyscale intensity is set by the logarithm of the (order-unity) frequency diffusion time. The darker pixels have shorter frequency diffusion times and are more chaotic. This map reveals the intersection of this particular energy hypersurface with the rich structure of resonant surfaces present in this potential and highlights the accuracy of \superfreq\ --- weak chaos (grey) is detectable over just 10's of orbital periods, compared to the many thousands of orbits it would take to detect such features with the maximum Lyapunov exponent. The tube orbits in this potential are mostly regular or only mildly chaotic --- the largest regular regions are associated with the short-axis and long-axis tube orbits --- however islands of stronger chaos do appear, especially at the intersections of resonances where resonance overlap occurs.

The strongest chaotic regions (black) appear in both of the above grids (Figure~\ref{fig:lyapmap} and Figure~\ref{fig:freqdiff}). Some of the weakly chaotic unstable resonances do appear in the Lyapunov time map, especially where resonances overlap --- for example, near $(x,z) \approx(13,7)$ and $(x,z) \approx(31,13)$ where there is a slight hint of weak chaos (grey) in Figure~\ref{fig:lyapmap}. The details of the resonant structure is revealed in the frequency diffusion map from integrations of just 80 orbital periods, though the majority of the orbits have estimated chaotic timescales corresponding to thousands of orbital periods. In the next section, we analyze the density evolution of finite-volume ensembles of orbits around each orbit in the above grids and compare the density evolution of the ensembles to the single-orbit chaos indicators computed in this section.

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant \emph{local} to a given orbit --- that is, these quantities are measures of how infinitesimal deviations will diverge on average from some parent orbit of interest, or of how long it takes for the frequencies of a single orbit to change by some amount. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy \citep[assuming masses of $10^4$--$10^5$~\msun;][]{johnston98}, but for a dwarf-galaxy-scale progenitor, the dispersion can be $\gtrsim$10\%. Does the Lyapunov or frequency diffusion time predict the timescale over which a finite phase-space volume (e.g., tidal debris) stays coherent?

We start by considering three particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular orbit (R), a weakly chaotic orbit (W), and a strongly chaotic orbit (S). The orbits were chosen from a single column of the orbit grid based on their frequency diffusion time to sample a range of possible values, and the initial conditions and chaos diagnostics are listed in Table~\ref{tbl:three-orbits}. We use the Self-Consistent Field (SCF) basis function expansion code \citep{hernquist92} to run N-body simulations of globular-cluster-mass progenitor systems on these three orbits. The initial mass of the clusters is \todo{kvj}{2.5E4 \msun contained within what again? 35 scale radii? but this is a plummer sphere?}. For each simulation, the progenitor is initialized as a $10^4$ particle Plummer sphere and the background triaxial NFW potential is turned on slowly over 250 Myr to reduce artificial gravitational shocking. We start the progenitor systems at apocenter and evolve for $\approx$16 orbital periods so that each simulation finishes again at subsequent apocenter. Figure~\ref{fig:nbodysims} shows the final particle distributions rotated so that the angular momentum of the progenitor orbit is aligned with the $z$-axis. A thin stream forms on the regular orbit (left column), a more diffuse stream on the mildly chaotic orbit (middle column), and a `fanned' stream on the strongly chaotic orbit (right column). Given the long Lyapunov and frequency diffusion times of the mildly chaotic orbit, it is surprising that stream formed on this orbit appears to be more diffuse than the regular orbit stream.

>>>>

To explore the question of if and how chaos manifests in the density evolution of orbit ensembles over timescales much shorter than that predicted from generic chaos indicators, we must first understand the behavior of individual orbits in complex gravitational potentials and the orbital structures in the potential itself (i.e. the strength of resonances and chaos). An orbit in an $N$ degree of freedom (dof) Hamiltonian, $H$, is a set of $2N$ quasi-periodic time series, 
\begin{equation}
(w_1(t),...,w_{2N}(t)) = (q_1(t),...,q_{N}(t),p_1(t),...,p_{N}(t)) \label{eq:coords}
\end{equation}
where $q_i$ and $p_i$ are conjugate coordinates in the sense that
\begin{align}
	\dot{p}_i &= -\frac{\partial H}{\partial q_i}\\
	\dot{q}_i &= \frac{\partial H}{\partial p_i}
\end{align}
for all $t$. If bounded, the motion in any component, $w_i(t)$, can be described as a Fourier sum,
\begin{equation}
	w_i(t) = \sum_k^\infty a_k \, e^{i\,\omega_k\,t} \label{eq:fourier}
\end{equation}
where the $a_k$ are complex amplitudes.

@@

A Hamiltonian or potential is said to be \emph{globally integrable} when the number of isolating integrals of motion is equal to the number of degrees of freedom and a transformation to angle-action coordinates may be done globally --- for example, the transformation to angle-action coordinates may be written as a function of arbitrary phase-space coordinates and the functional form is independent of phase-space position \citep[e.g.,][]{goldstein80}. The condition for global integrability is very restrictive and it seems that the likelihood that a Hamiltonian is globally integrable decreases as the number of degrees of freedom increase \citep[e.g.,][]{lichtenberg83}. In a globally integrable potential, the Hamiltonian may be written solely in terms of the actions, $H = H(\boldsymbol{J})$. Galactic potentials are almost certainly not globally integrable, but it is useful to understand the orbit structure in integrable systems before extending to more general potentials. Only one globally integrable triaxial mass model is known that generates motion that is qualitatively similar to that seen in Galaxies: the St\"ackel or `Perfect Ellipsoid' model \citep[e.g.,][]{kuzmin73, deZeeuw85}. The four general classes of orbits in this potential are box, inner long-axis tube, outer long-axis tube, and short-axis tube orbits. The tube orbits preserve a sense of rotation about either the long or short axis and are therefore centrophobic, whereas the box orbits may approach arbitrarily close to the center of the potential (centrophilic) and may change sense of rotation about any axis.

The frequencies of a generic orbit are typically incommensurable --- that is, $\bs{n} \cdot \bs{\Omega} \neq 0$ for any integer vector, $\bs{n}$, with reasonable magnitude.\footnote{A more precise condition is stated in terms of a diophantine condition, e.g., $|\bs{n} \cdot \boldsymbol{\Omega}| > \alpha \, |n|^{-\gamma}$ where $\alpha, \gamma>0$.} These conditionally-periodic orbits uniformly cover the surface of an orbital torus. If instead there exists a relation of the form $\boldsymbol{n} \cdot \boldsymbol{\Omega} = 0$ the orbit is referred to as a resonant orbit. Resonant orbits are confined to a surface with lower dimensionality than the surface of an orbital torus, depending on the number of resonance relations obeyed: we refer to orbits that obey a single resonance relation as \emph{uni-resonant} orbits, and if an additional resonance relation exists, \emph{bi-resonant}. Uni-resonant orbits in a triaxial potential are confined to a 2D surface, and bi-resonant orbits are closed 1D curves. The resonant structure of a potential --- the relative importance of particular resonance numbers --- is difficult to compute, but determines the global behavior of orbits in the potential. For a triaxial potential the resonances define planes in frequency-space, but resonance surfaces in action-space are generic surfaces. In plots of frequency ratios, the resonances appear as lines; Figure~\ref{fig:cartoons} (left panel) shows a cartoon portrait of a portion of frequency-space for an integrable potential with example resonance lines, non-resonant, and resonant orbits marked. For an integrable potential, all orbits are regular.

@@

The orbit structure of near-integrable potentials can be understood by considering a Hamiltonian that is a small perturbation away from being globally integrable --- that is, a Hamiltonian that may be written
\begin{equation}
	H(\boldsymbol{J}, \boldsymbol{\theta}) = H_0(\boldsymbol{J}) + \epsilon \, H_1(\boldsymbol{J}, \boldsymbol{\theta})
\end{equation}
where $H_0(\bs{J})$ represents an integrable Hamiltonian, and $\epsilon$ is a small parameter that determines the perturbation strength \citep[a description of perturbation theory applied to nonlinear Hamiltonians is given in][]{lichtenberg83}. When $0 < |\epsilon| \ll 1$, resonant surfaces become `thick' resonant layers, within which orbits are qualitatively similar to the parent resonant orbit \citep[e.g.,][]{merritt99}. These resonant layers are then generically surrounded by stochastic layers where chaotic motion occurs (in the vicinity of separatrices). Chaotic orbits are characterized by exhibiting random behavior despite being generated by entirely deterministic equations of motion. Actions strictly do not exist for chaotic orbits and therefore chaos `destroys' orbital tori in certain regions of action-space --- these frequency spectrum of orbits in these regions evolve chaotically with time.

Figure~\ref{fig:cartoons} (middle panel) shows a cartoon of frequency space for a near-integrable potential (a small perturbation away from the potential in the left panel). Much of the structure that was present in the integrable potential remains in the near-integrable case, but the differences are highlighted. Orbits in the resonant layers surrounding the resonances (grey) are near-resonant orbits that librate around the resonance. Chaotic orbits in the stochastic layer (red) behave erratically depending on the surrounding resonance structure. If the stochastic layer is small and the chaotic orbit is therefore confined, the orbit may behave nearly regular for long periods of time.

For small values of $\epsilon$, many regular orbits survive and only small chaotic regions are introduced, especially where resonances overlap \citep[see][]{chirikov60}. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Figure~\ref{fig:cartoons} (right panel) qualitatively shows this phenomenon --- the resonant and resonant-layer orbits may still be regular, but many or all of the conditionally-periodic orbits can be chaotic. As the perturbation strength increases, next the uni-resonant tori are destroyed, and finally the bi-resonant tori --- these are least susceptible to destruction from perturbations \cite[for a more quantitative illustration of this transition from integrability to global chaos, see Figure~9 in][]{valluri98}.

When $\epsilon$ is large, there is no general prediction for the resulting behavior, however it seems that more complicated and physically motivated triaxial potential models for galaxies follow the intuition gained from the small-perturbation picture described above, at least for certain parameter choices \citep[e.g.,][]{valluri98, merritt99}. We therefore expect a large number of regular orbits will survive --- the so-called KAM tori --- however the tori that survive will be separated by regions of chaotic motion. Any transformations to angle-action coordinates must be defined local to each resonance region due to the destruction of tori and chaotic motion which lead to discontinuous changes in orbital properties.

@@

Chaotic orbits have no orbital actions because they are only strictly bound to their energy hypersurface (if the potential is time-independent). The orbits therefore do not have a single set of fundamental frequencies, but rather the frequencies that describe the character of motion evolve with time. In near-integrable potentials, chaotic diffusion of an orbit in actions and frequencies occurs both across resonance layers (a sort of stochastic libration) and along the stochastic layers that surround the resonances (Arnold diffusion). For weakly chaotic orbits, motion across a resonance layer can occur with a frequency close to the libration frequency of the nearby stable orbits in the resonance layer. Thus, if the resonance libration frequency is comparable (within a factor of a few) to the orbital frequency, motion across a resonance can modulate the frequency spectrum of an orbit over an orbital time. However, the stochastic layers are often bounded in the direction orthogonal to the resonance by other stable, resonant regions so that the frequencies or actions can not change by large factors (unless there is resonance overlap and the motion is strongly chaotic).

The rate of diffusion along stochastic layers via Arnold diffusion depends on the local resonant structure and is hard to predict. This has been done analytically for simple potentials \citep[e.g.,][]{chirikov79}. For systems with $N>2$ dof, the stochastic layers connect and form an intricate network of stochasticity known as the Arnold web; an orbit that ergodically mixes over its energy hypersurface must traverse this web, though the timescales typical for this phenomenon are many thousands of orbital periods.

Arnold diffusion is not expected to be significant for most orbits over timescales relevant to galaxies (10s of orbits), however chaotic motion across resonances can occur over short times. If a stochastic orbit is semi-bounded by surrounding resonances, the allowed volume in which it can explore (in frequency space) is often quite small so that the orbit may appear to be approximately regular. However, this small-scale evolution may be important for tidal debris which can have small intrinsic spreads in frequencies.

\subsection{Mixing of orbit ensembles}\label{sec:chaotic-mixing}

An ensemble of regular orbits (e.g., tidal debris) will phase-mix due to (small) differences in the fundamental frequencies of the orbits. Generically, a small ensemble of chaotic orbits will lose coherence much faster than for regular orbits \cite[see, e.g.,][]{kandrup94, merritt96, kandrup03}, however this depends on the details of the resonant structure around the ensemble and are thus difficult to predict. For example,  ensembles of orbits `stuck' between resonances may quickly spread to fill the allowed volume via the short-time chaotic diffusion processes, but then the orbits must escape this confinement and diffuse through the Arnold web to reach a fully mixed state \citep{merritt96}.

In this work, we investigate the consequences of this short-time but small-scale frequency evolution and hypothesize that this may explain the enhanced density evolution of tidal debris around weakly chaotic regions where chaotic timescales are predicted to be long \citep[e.g.,][]{pearson15}. We then discuss how this would affect our understanding of the coherence and density evolution of tidal streams.

@@

Our goal is to map the orbit structure of arbitrary (galactic) potentials, with an emphasis on identifying the chaotic orbits and understanding the evolution of these ensembles of orbits over short times. In particular, we aim to understand how this chaos-enhanced density evolution can affect tidal stream morphology. In this section, we describe the methods we will use to detect and quantify the strength of chaos for large grids of orbits.

@@

\begin{table*}[ht]
\begin{center}
	\begin{tabular}{ c  c }
	         Parameter & Value\\\toprule
		$v_c$ & 175~km~s$^{-1}$\\
		$r_s$ & 20~kpc\\
		$a$ & 1\\
		$b$ & 0.77\\
		$c$ & 0.55\\
		\bottomrule
		\end{tabular}
	\caption{Summary of parameters for the triaxial NFW potential (Equation~\ref{eq:potential}) used in this work. Triaxiality is introduced in the density (rather than the potential) to ensure that the density is physical at all radii. Velocity scale, scale radius, and axis ratios are chosen to match the median halo parameters for a $M_{\rm vir} \approx 10^{12}~\msun$ halo from \citep{jing02}. \label{tbl:potential}}
\end{center}
\end{table*}

@@

If an orbit is chaotic, the motion can no longer be expanded in terms of a single set of fundamental frequencies. For a weakly chaotic orbit, the orbit may appear consistently periodic over long windows of time. \superfreq\ will pick out a set of frequencies for chaotic orbits that correspond to the largest peaks in the power spectrum of the orbits, however these peaks will change location and amplitude with time. For more strongly chaotic orbits, the power spectrum will be quite noisy and the peak frequencies may change erratically when comparing two consecutive sections of orbit. The frequencies picked out by \superfreq\ for such orbits will therefore represent the average periodic nature of the orbit over a given integration window. We define the fractional frequency diffusion rate per orbital period, $\mathcal{R}$, in the $k$th fundamental frequency as
\begin{equation}
	\mathcal{R}_k = \frac{\Omega_{k}^{(2)} - \Omega_{k}^{(1)}}{\Omega_{k}^{(1)} \, N} \label{eq:fdrate}
\end{equation}
where the upper index refers to the two consecutive sections of orbit and $N$ is the number of orbital periods in a single section of orbit. By inverting this rate, we can compute the timescale over which we expect order-unity changes to the fundamental frequencies: the \emph{frequency diffusion time} is defined as
\begin{equation}
	t_\Omega = \, (\max_{a_k} \, \mathcal{R}_k)^{-1} \label{eq:fdtime}
\end{equation}
where the maximum is taken with respect to the corresponding amplitudes, $a_k$, of the fundamental frequency components \citep[see][]{valluri12}, and the time is in units of number of orbital periods, ${\rm T}$.

@@

In Section~\ref{sec:results1}, we generate grids of orbits in the potential described above to map the orbit structure of the potential. We classify each orbit in terms of the strength of chaos along the orbit as computed using the Lyapnov and frequency diffusion times of Section~\ref{sec:methods}. With this initial classification, in Section~\ref{sec:results2} we follow the evolution of ensembles of orbits generated around each orbit in the initial grid. We find that the configuration-space density of ensembles around weakly chaotic orbits evolve faster (e.g., in mean density) than expected given the timescales over which chaos is computed to be relevant for the parent orbits. In Section~\ref{sec:results3}, we explain this phenomenon in the context of how chaotic diffusion occurs (e.g., Section~\ref{sec:behavior-chaotic}).

@@

We generate an isoenergy grid of initial conditions along the $xz$ ($y=0$) plane\footnote{$x$ is the major and $z$ the minor axis.} with energy (per unit mass), $E$, chosen to span a range of distances comparable to the scale radius of the potential ($E=-(397.2~{\rm km~s}^{-1})^2$ in physical units; see Table~\ref{tbl:potential}). We fix $v_x = v_z = 0$, and compute $v_y$ from the energy. The majority of the orbits generated on this grid are tube orbits, which preserve a sense of rotation about one of the principal axes of the potential and are generally centrophobic. Tube orbits are generally less stochastic than box orbits, which tend to plunge deep into the inner regions of the potential where the density profile is steep, however there are appreciable numbers of stochastic tube orbits. The major classes of stable tube orbits circulate about either the major or minor axis. This grid generates all of the major orbit classes (short-axis tubes, inner long-axis tubes, outer long-axis tubes, stochastic intermediate-axis, and box orbits), but the most numerous orbits in this grid are the short-axis and outer long-axis tubes. Thin tidal streams are thought to form along tube orbits rather than box orbits because of the faster disruption expected for stellar systems on radially plunging orbits which makes them more diffuse.

\todo{apw}{New figure showing just the grid and orbit classifications, labels for orbit types}

@@

For each orbit, we also separately integrate for $\approx$80 orbital periods and use \superfreq\ to compute the fundamental frequencies for the two consecutive sections of 40 orbital periods. We have found that this is the shortest integration window for which we are able to successfully recover frequencies using \superfreq\ for $>$99.9\% of the orbits. Figure~\ref{fig:freqdiff} shows the same grid of initial conditions as in Figure~\ref{fig:lyapmap}, but now the greyscale intensity is set by the logarithm of the (order-unity) frequency diffusion time. The darker pixels have shorter frequency diffusion times and are more chaotic. This map reveals the intersection of this particular energy hypersurface with the rich structure of resonant surfaces present in this potential and highlights the accuracy of \superfreq\ --- weak chaos (grey) is detectable over just 10's of orbital periods, compared to the many thousands of orbits it would take to detect such features with the maximum Lyapunov exponent. The tube orbits in this potential are mostly regular or only mildly chaotic --- the largest regular regions are associated with the short-axis and long-axis tube orbits --- however islands of stronger chaos do appear, especially at the intersections of resonances where resonance overlap occurs.

The strongest chaotic regions (black) appear in both of the above grids (Figure~\ref{fig:lyapmap} and Figure~\ref{fig:freqdiff}). Some of the weakly chaotic unstable resonances do appear in the Lyapunov time map, especially where resonances overlap --- for example, near $(x,z) \approx(13,7)~{\rm kpc}$ and $(x,z) \approx(31,13)~{\rm kpc}$ where there is a slight hint of weak chaos (grey) in Figure~\ref{fig:lyapmap}. The details of the resonant structure is revealed in the frequency diffusion map from integrations of just 80 orbital periods.

While there is rich structure and a significant number of weakly chaotic orbits, the majority of the orbits have estimated chaotic timescales corresponding to thousands of orbital periods and are thus not expected to be relevant for tidal stream evolution. In the next section, we analyze the density evolution of finite-volume ensembles of orbits around each orbit in the above grids in order to study the effect of weak chaos on tidal debris. We then compare the density evolution of the ensembles to the single-orbit chaos indicators computed in this section.

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant for a given orbit --- that is, these quantities are measures of how infinitesimal deviations will diverge on average from some parent orbit of interest, or of how long it takes for the frequencies of a single orbit to change by some amount. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy \citep[assuming masses of $10^4$--$10^5$~\msun;][]{johnston98}, but for a dwarf-galaxy-scale progenitor, the dispersion can be $\gtrsim$10\%. In this section, we ask whether  the Lyapunov or frequency diffusion time predict the timescale over which a finite phase-space volume (e.g., tidal debris) stays coherent.
